---
title: 'Regression Analysis: MPG in Automatic vs Manual cars'
author: "Edwin Seah"
date: "19 May 2015"
output:
  pdf_document: default
  html_document: default
fontsize: 10pt
---
```{r global_options, include=FALSE}
# Setting global options for R markdown
knitr::opts_chunk$set(fig.path='figures/',
                      echo=TRUE, 
                      warning=FALSE,
                      message=FALSE)
```

### Executive Summary

This analysis checks if miles per gallon (MPG) benefit more from automatic versus manual transmissions, and quantifies any such difference. Although there is a stark difference in expected MPG between automatic and manual cars, in itself it is not a realistic predictor since the variables **wt** (lb/1000) and **qsec** (1/4 mile time) have significant influences on MPG, settling finally on our stepwise derived model of **mpg ~ factor(am) + wt + qsec**, after checking it versus a designed model involving groups of regressors that mpg is likely dependent on.

### Getting/Transforming Data and some Exploratory Data Analysis  

The **mtcars** dataset comprises fuel consumption (MPG) and 10 aspects of automobile design and performance for 32 cars, loaded as `data(mtcars)` and stored in a data frame **m**. We transform **am** as factor variable of 2 levels ("Automatic, "Manual"). Cursorily our box-whisker plot (Fig.1) indicates Manual transmissions have a clear advantage over Automatic transmissions in MPG terms.
```{r load_data, echo=FALSE, cache=TRUE}
data(mtcars) ; m <- mtcars ; 
m$am <- factor(m$am, labels=c("(Automatic)", "(Manual)"))
#m$cyl <- factor(m$cyl, labels=c(4, 6, 8))           # not used
#m$vs <- factor(m$vs, labels=c("(V)", "(Straight)")) # not used
#m$carb <- factor(m$carb, labels=c(1,2,3,4,6,8))     # not used
```

### Quantifying the relationship via Regression Analysis  

As a baseline we simply fit **mpg** (outcome) against Transmission Type **am** (predictor). 
```{r lm_mpg_trans, cache=TRUE}
m$am <- factor(m$am, labels=c("(Automatic)", "(Manual)"))
fitam <- lm(mpg ~ am, data=m) ; summary(fitam)$coef
```
This model gives an MPG expected gain of **`r round(summary(fitam)$coef[2], 2)`** going from an Automatic to Manual transmission. However, our adjusted R^2^ is **`r round(summary(fitam)$adj.r.squared, 2)`** (DF = `r fitam$df[2]`); low as we had not fit 10 other candidate regressors. Residuals (Fig.2) exhibit homoskedacity (evenly scattered around 0) and nearly normally distributed, but only **`r round(summary(fitam)$adj.r.squared*100, 2)`%** of MPG variability was explained.

Given limitations in explaining MPG variability with just **am**, a quick parsimonious model can be found using a mechanical backwards stepwise elimination approach (at a somewhat abritrary siginificance level of ($\alpha$ = 5%). For code brevity we use the inbuilt automated AIC method by calling`step()` (*fstep*); it gives us the same resultant model as the manual way (*fman*, see Fig. 3).
```{r lm_mpg_backwards_elimination_manual, echo=FALSE, cache=TRUE}
fitb1 <- lm(data=m, mpg ~ .)
fitb2 <- update(fitb1, mpg ~ . - cyl)
fitb3 <- update(fitb2, mpg ~ . - cyl - vs)
fitb4 <- update(fitb3, mpg ~ . - cyl - vs - carb)
fitb5 <- update(fitb4, mpg ~ . - cyl - vs - carb - gear)
fitb6 <- update(fitb5, mpg ~ . - cyl - vs - carb - gear - drat)
fitb7 <- update(fitb6, mpg ~ . - cyl - vs - carb - gear - drat - disp)
fman <- update(fitb7, mpg ~ . - cyl - vs - carb - gear - drat - disp - hp)
```
```{r lm_mpg_backwards_elimination_auto}
full <- lm(data=m, mpg ~ .) ; fstep <- summary(step(full, direction="backward", trace=0))
print(rbind(fman$coef, fstep$coef[1:4]))
```
We arrive at a model with an adjusted R^2^ of **`r round(summary(fman)$adj.r.squared, 2)`** (DF = `r fman$df[2]`), with residuals showing some tail-skew in the normal probability plot (Fig. 4).

To check inflation of the estimate's variance by regressor groups we design a model that includes suspected/likely dependent variables of **mpg**, fitting the following models of interest (groups) in order:

Group      | Weight | Engine Power | Engine Configuration | Gearing
-----------|--------|--------------|----------------------|--------
Regressors | wt | disp, hp | cyl, carb, vs | gear, drat
Model | fit1 | fit2 | fit3 | fit4

Using a nested liklihood ratio test (*fit1* to *fit4*) with our base *fit* helps check their contribution to **mpg** via the ANOVA results:
```{r lm_nested_model_testing_fit, echo=FALSE}
fit <- lm(mpg ~ am, data=m)
fit1 <- update(fit, mpg ~ am + wt)
fit2 <- update(fit, mpg ~ am + wt + disp + hp)
fit3 <- update(fit, mpg ~ am + wt + disp + hp + cyl + carb + vs)
fit4 <- update(fit, mpg ~ am + wt + disp + hp + cyl + carb + vs + gear + drat)
fit5 <- lm(mpg ~ am + wt + qsec, data=m)
```
```{r lm_nested_model_testing_fit_anova}
anova(fit, fit1, fit2, fit3, fit4)[1:6]
```
```{r cov_check}
cv <- function(f) {summary(f)$cov.unscaled[2,2]}
c(cv(fit1), cv(fit2), cv(fit3), cv(fit4), cv(fit5))/cv(fit)
```
```{r finalfit, echo=FALSE}
s2 <- summary(fit2) ; s5 <- summary(fit5)
```
We would opt for *fit2* (**p-value = `r round(anova(fit, fit1, fit2, fit3, fit4)[1:6][3,6],4)`**), rejecting for lack of significance) over the others. Our covariances and adjusted R^2^ for *fit2* (**`r round(cv(fit2)/cv(fit), 2)`**, **`r round(s2$adj.r.squared, 2)`**) and *fit5* (**`r round(cv(fit5)/cv(fit), 2)`**, **`r round(s5$adj.r.squared, 2)`**) are similar, with *fit2* residuals (Fig. 5) showing the Maserati Bora exerting very high leverage. VIF for *fit2* regressors are higher than in our stepwise model *fit5*:
```{r vif_check}
library(car) ; sqrt(vif(fit2)) ; sqrt(vif(fit5))
```
We note that quarter mile time **qsec** has a very low VIF viz both **hp** and **disp**, which are likely colinear. Intuitively **qsec** may be a good proxy for any/all of the engine power/configuration variables. We conclude in favour of *fit5* (**mpg ~ factor(am) + wt + qsec**); it is simpler (one less regressor) than *fit2* with a marginally better adjusted R^2^ of **`r round(s5$adj.r.squared, 2)`**, giving an expected **`r round(s5$coef[3], 2)`** MPG per 1000lbs increase in weight and **`r round(s5$coef[2], 2)`** gain going to a Manual transmission, and **`r round(s5$coef[4], 2)`** MPG gain per 1 second slower 1/4 mile timing **qsec**.

### Project Repo

+ All files and full code used are available from the [Github Project Repository](https://github.com/slothdev/RMproject-Repo) (https://github.com/slothdev/RMproject-Repo)  

### Appendix

### Fig. 1 - MPG by Transmission Type

Both the median and inter-quartile range (or middle 50% of all cars) for Manual transmission type cars are clearly higher than Automatic transmission cars.
```{r fig1_eda_mpg_trans}
boxplot(mpg ~ am, 
        data=m, 
        xlab="Transmission Type", 
        ylab="Miles per Gallon", 
        names=c("Automatic", "Manual"), 
        main = "Fig.1 - MPG by Transmission Type")
```

### Fig. 2 - Residual and QQ plots of MPG by Transmission Type

```{r fig2_qq_resid_mpgByTrans}
par(mfrow=c(1,2))
# Residuals plot
plot(resid(fit), main="Residual Plot (mpg ~ am)")
abline(a=0, b=0)
# Normal Probability Plot
qqnorm(rstandard(fit), 
       ylab="Standardized Residuals", 
       xlab="Normal Scores")
qqline(rstandard(fit))
```

### Fig. 3 - Simple backwards elimination stepwise by highest p-value

1. Start with a full model, as it provides an unbiased variance estimate for MPG due to including all variables. It may contain regressors with high colinearity and litle unique contribution to **mpg**.

2. Eliminate one regressor variable at a time (whichever has the highest p-value from the T-test) and refit.

3. Stop eliminating when no regressor has a p-value higher than $\alpha$ or when our adjusted R^2^ stops going up.

These intermediate steps proceed as follows:
```{r fig3_lm_mpg_backwards_elimination_summary}
showp <- function(b) {summary(b)$coeff[,4]}
showp(fitb1)
showp(fitb2)
showp(fitb3)
showp(fitb4)
showp(fitb5)
showp(fitb6)
showp(fitb7)
showp(fman)
```

\newpage

### Fig. 4 - Residuals from backwards elimination (fman aka fit5)

```{r fig4_lm_mpg_backwards_elimination_resplot, echo=FALSE, eval=FALSE}
par(mfrow=c(1,2))
# Residuals plot
plot(resid(fman), main="Residual Plot (MPG ~ am + wt + qsec)")
abline(a=0, b=0)
# Normal Probability Plot
qqnorm(rstandard(fman), ylab="Standardized Residuals", xlab="Normal Scores")
qqline(rstandard(fman))
```
```{r fig4_lm_fman_plotfit, fig.align='center'}
par(mfrow=c(2,2))
plot(fman)
```

\newpage

### Fig. 5 - Residuals from fit2

```{r fig5_lm_fit2_plotfit, fig.align='center'}
par(mfrow=c(2,2))
plot(fit2)
```

```{r echo=FALSE, eval=FALSE}
newdata = data.frame(am=mean(am))
predict(mpgByWt.lm, data.frame(am=mean(am), interval="confidence"))
```
